---
title: "Getting Started with pTRAPPING"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pTRAPPING}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%",
  dpi = 150,
  message = FALSE,
  warning = FALSE
)
```

## What problem does pTRAPPING solve?

Imagine you want to know which genes are actively being read — translated into
protein — in a specific, rare cell type buried inside a complex tissue like the
brain. You cannot easily sort those cells out; they are intermingled with
hundreds of other cell types. TRAP-seq and PhosphoTRAP are molecular techniques
that let you tag ribosomes (the cellular machinery that reads genes) in your
target cells, and then pull them out with their attached RNA. The RNA you
capture in this immunoprecipitation step — the **IP fraction** — reflects what
your target cells were actively translating. You keep the leftover, un-pulled
RNA too — the **INPUT fraction** — as a reference for total gene expression in
the tissue.

The core question is then: **which genes are enriched in the IP relative to the
input?** A gene enriched in IP is being translated in your target population; a
gene depleted in IP is present in the tissue but not in your cells of interest.
Answering this question requires normalising the data, picking a statistical
test, adjusting for multiple comparisons across thousands of genes, and
visualising the results — all of which pTRAPPING handles for you with three
functions.

**PhosphoTRAP** adds one more layer: instead of constitutively tagging
ribosomes, it captures only ribosomes that carry a phosphorylated form of the
S6 protein — a marker of recent neural activation. This means you can ask not
just "which genes are translated in these cells?" but "which genes are
translated in these cells *right now*, after a specific stimulus?". The
experimental design is otherwise identical: IP fraction (activated cells'
RNA) vs. INPUT fraction (total tissue RNA).

## The example dataset: Tan et al. (2016)

Throughout this vignette we use data from
[Tan et al. (2016) *Cell* 167, 47–59](https://doi.org/10.1016/j.cell.2016.08.028),
a landmark PhosphoTRAP study that identified **warm-sensitive neurons** in the
mouse preoptic area (POA) of the hypothalamus — the brain region that controls
body temperature.

Mice were briefly exposed to warmth (37 °C). Neurons that fired in response
had their ribosomes phosphorylated, which were then isolated by
immunoprecipitation. Two groups of mice were compared:

* **PACAP** — warm-exposed mice (their warm-activated neurons were captured)
* **BDNF** — baseline control mice

Each group has **2 biological replicates** (2 mice), giving 8 samples total:
4 for PACAP (2 IP + 2 INPUT) and 4 for BDNF (2 IP + 2 INPUT).

> **A note on replicates:** n = 2 is small for GLM-based methods. For this
> dataset we use the **paired t-test**, which is the approach from the original
> paper. For your own experiments, aim for ≥ 3–4 replicates; with ≥ 4,
> `test_method = "LRT"` is the recommended default.

```{r load-data}
library(pTRAPPING)

# Raw integer counts — use with GLM methods or CPM/mratios normalisation
counts_raw <- read.delim(
  system.file("extdata", "TAN_etal_2016_raw.txt", package = "pTRAPPING")
)

# Pre-computed RPKM values — already normalised; use with norm.method = "none"
counts_rpkm <- read.delim(
  system.file("extdata", "TAN_etal_2016_RPKM.txt", package = "pTRAPPING")
)

dim(counts_raw)   # 8862 genes × 9 columns (1 gene-ID + 8 samples)
names(counts_raw) # column names encode treatment, fraction, and replicate
```

The package ships two versions of the same dataset:

* **`TAN_etal_2016_raw.txt`** — raw integer read counts, the starting point
  for GLM-based methods (`"LRT"`, `"QLF"`, `"deseq"`, `"voom"`) and for
  `"CPM"` or `"mratios"` normalisation inside the t-test branch.
* **`TAN_etal_2016_RPKM.txt`** — the same data already normalised to RPKM
  by the original authors. Because the values are pre-normalised, use
  `norm.method = "none"` when passing this matrix — this is the approach taken
  in Tan et al. (2016) and is what we demonstrate in the worked examples below.

The column names follow the convention pTRAPPING expects: each name contains a
**treatment label** (`PACAP` or `BDNF`), a **fraction keyword** (`IP` or
`Input`), and a **replicate number** (`1` or `2`), in any order and separated
by underscores. The function parses these automatically — no metadata table
needed.

---

## Step 1 — Differential expression with `ptrap_de()`

`ptrap_de()` compares the IP fraction to the INPUT fraction for a given
treatment group and returns a ranked table of genes. The key number it produces
is the **log2 fold change (logFC)**:

* **logFC > 0** → the gene is *more abundant* in IP than INPUT (enriched in
  your target cells)
* **logFC < 0** → the gene is *less abundant* in IP (present in the tissue but
  not your target cells)
* A logFC of **1** means the gene is exactly **2× more abundant** in IP;
  a logFC of **2** means **4×**, and so on.

### Choosing a statistical method

Six methods are available. Use this table as a quick guide:

| Method | Package | Min. replicates | Notes |
|---|---|:---:|---|
| `"LRT"` | edgeR | 4 | **Recommended default** for most experiments |
| `"QLF"` | edgeR | 4 | More conservative than LRT |
| `"deseq"` | DESeq2 | 4 | Use `shrink.lfc = TRUE` for figures/ranking |
| `"voom"` | limma | 4 | Good for complex designs with covariates |
| `"paired.ttest"` | base R | 3 | Per-animal IP/INPUT pairing; small n |
| `"unpaired.ttest"` | base R | 4 | Compares enrichment *between* two groups |

All four GLM-based methods (`"LRT"`, `"QLF"`, `"deseq"`, `"voom"`) normalise
the data internally — you do not need to normalise first. For the t-test
methods you can choose a normalisation via `norm.method`:

| `norm.method` | When to use |
|---|---|
| `"CPM"` (default) | Raw counts; comparable across replicates of the same gene |
| `"RPKM"` | Raw counts + gene lengths available via `gene.length`; needed for cross-gene comparisons within a sample |
| `"mratios"` | Raw counts; DESeq2-style size-factor normalisation |
| `"none"` | Data already normalised (e.g., a pre-computed RPKM matrix) |

> **Note on `"RPKM"`:** this option requires you to supply gene lengths via
> `gene.length` (a named numeric vector, in base pairs). Without it, the
> function will throw an error. Gene lengths can be obtained from annotation
> packages such as `TxDb.*` or via BioMart. When you already have a
> pre-normalised matrix — as in the Tan et al. dataset below — use
> `norm.method = "none"` instead.
>
> **Note on `filter = FALSE`:** by default, `ptrap_de()` removes low-expression
> genes using `filterByExpr`, which expects raw integer counts. When passing a
> pre-normalised matrix (RPKM, TPM, or similar), the small floating-point values
> will mislead the filter and cause real genes to be dropped. Set `filter = FALSE`
> whenever `norm.method = "none"` is used with a pre-normalised input.

### Running the paired t-test (Tan et al. approach)

Because this dataset has only n = 2 replicates per group, we use the paired
t-test — the same approach as the original paper. Tan et al. ran the test on
RPKM values, so we pass the pre-normalised RPKM matrix and set
`norm.method = "none"` to skip any additional normalisation step.

```{r ptrap-de-ttest}
res_PACAP <- ptrap_de(
  counts_mat = counts_rpkm,
  treatment_name = "PACAP",
  test_method = "paired.ttest",
  norm.method = "none",
  input_level = "Input", # matches "Input" (not "INPUT") in the column names
  filter = FALSE          # skip count-based filtering; data are pre-normalised
)
```

The function prints a message showing what it auto-parsed from the column
names: treatments, replicate numbers, and fractions. Inspect the result:

```{r inspect-results}
# The paired.ttest method always returns a named list
names(res_PACAP)

# $results: one row per gene, sorted by p-value
head(res_PACAP$results[, c("Gene", "logFC", "PValue", "FDR", "diffexpressed")])
```

### Understanding the output columns

| Column | What it means |
|---|---|
| `Gene` | Gene identifier |
| `logFC` | log2(IP / INPUT) — positive = enriched in target cells |
| `PValue` | Raw p-value from the paired t-test |
| `FDR` | Benjamini-Hochberg adjusted p-value (false discovery rate) |
| `diffexpressed` | `"UP"` (enriched), `"DOWN"` (depleted), or `"NO"` |

The `diffexpressed` column is determined by two thresholds (both must be met):

* `lfc_threshold` — minimum absolute logFC (default `1`, meaning ≥ 2× change)
* `fdr_threshold` — maximum FDR (default `0.05`)

You can change these without re-running the analysis — `ptrap_volcano()` and
`ptrap_volcano2()` recompute the classification from your chosen thresholds.

### $fe: per-animal fold enrichments

The paired t-test also returns a `$fe` tibble with each animal's IP/INPUT fold
enrichment (FE) for every gene. This is useful for spot-checking individual
genes or running your own downstream analyses.

```{r inspect-fe}
# Per-animal fold enrichment for the top 5 genes
head(res_PACAP$fe, 5)
```

### Running with a GLM method (for better-powered experiments)

With ≥ 4 replicates, the edgeR LRT is the recommended method. The call is
nearly identical — just change `test_method` and drop `norm.method` (edgeR
normalises internally via TMM):

```{r ptrap-de-lrt, eval = FALSE}
# Example only — requires >= 4 replicates to produce reliable estimates.
# Use the raw count matrix (GLM methods normalise internally via TMM).
res_lrt <- ptrap_de(
  counts_mat = counts_raw,
  treatment_name = "PACAP",
  test_method = "LRT",
  input_level = "Input"
)
head(res_lrt[, c("Gene", "logFC", "PValue", "FDR", "diffexpressed")])
```

### Comparing enrichment *between* two conditions (unpaired t-test)

The `"unpaired.ttest"` method answers a different question: not "which genes
are enriched in IP vs. INPUT?" but "which genes show *different* enrichment
between the PACAP and BDNF groups?" Genes that are enriched equally in both
will score near zero; genes that are specifically enriched in warm-activated
cells will stand out.

```{r ptrap-de-unpaired, eval = FALSE}
res_compare <- ptrap_de(
  counts_mat = counts_rpkm,
  treatment_name = "PACAP",
  control_name = "BDNF",
  test_method = "unpaired.ttest",
  norm.method = "none",
  input_level = "Input"
)
head(res_compare[, c("Gene", "logFC", "diff_FE", "PValue", "FDR")])
```

The output adds `diff_FE` (the ratio of mean fold enrichments) and per-group
mean fold enrichments alongside the usual logFC, PValue, and FDR columns.

---

## Step 2 — Visualise with `ptrap_volcano()`

A **volcano plot** puts logFC on the x-axis and statistical significance
(-log p-value) on the y-axis. Genes in the **upper-right corner** are both
strongly enriched *and* statistically reliable — the most interesting
candidates. Genes in the upper-left are strongly depleted. Grey points fail
one or both thresholds.

```{r ptrap-volcano, fig.cap = "Volcano plot for the PACAP (warm-exposed) group. Known warm-sensitive markers are labeled."}
ptrap_volcano(
  res_PACAP$results,
  fdr = FALSE, # use raw p-values (n=2; FDR is very conservative)
  log_base = 2, # -log2(p) on y-axis, matching Tan et al. 2016
  genes.annot = c("Adcyap1", "Bdnf", "Ucn3", "Gng8"),
  title = "Warm-exposed neurons (PACAP) — Preoptic Area"
)
```

Key arguments:

| Argument | Default | What it does |
|---|---|---|
| `fdr` | `TRUE` | Use FDR (`TRUE`) or raw p-value (`FALSE`) for the y-axis and colouring |
| `lfc_threshold` | `1` | logFC cutoff for the vertical dashed lines |
| `fdr_threshold` | `0.05` | p-value cutoff for the horizontal dashed line |
| `log_base` | `10` | Base of the log transform on the y-axis |
| `genes.annot` | `NULL` | Character vector of gene names to label |
| `title` | auto | Plot title (defaults to region + treatment) |

### Customising colours

Pass a named vector to override the defaults:

```{r volcano-custom, eval = FALSE}
ptrap_volcano(
  res_PACAP$results,
  fdr = FALSE,
  colors = c("UP" = "#D55E00", "DOWN" = "#56B4E9"),
  title = "PACAP — custom colours"
)
```

The function returns a ggplot2 object, so you can add or override any layer
with standard ggplot2 syntax:

```{r volcano-ggplot, eval = FALSE}
library(ggplot2)

ptrap_volcano(res_PACAP$results, fdr = FALSE) +
  theme(plot.title = element_text(face = "bold")) +
  xlim(-10, 10)
```

---

## Step 3 — Compare two conditions with `ptrap_volcano2()`

When you have run `ptrap_de()` for two groups, `ptrap_volcano2()` places both
results on a single 2D scatter plot:

* **x-axis**: logFC for condition 2
* **y-axis**: logFC for condition 1
* **Diagonal (dotted)**: genes enriched equally in both
* **Above diagonal**: more enriched in condition 1
* **Below diagonal**: more enriched in condition 2

Genes that meet the significance thresholds in both conditions, in only one,
or in neither are shown in distinct colours.

```{r ptrap-de-bdnf}
# Run ptrap_de() for the second group using the same pre-normalised matrix
res_BDNF <- ptrap_de(
  counts_mat = counts_rpkm,
  treatment_name = "BDNF",
  test_method = "paired.ttest",
  norm.method = "none",
  input_level = "Input",
  filter = FALSE
)
```

```{r ptrap-volcano2, fig.cap = "Scatter comparison of PACAP and BDNF enrichment. Genes on the diagonal are enriched equally in both conditions."}
ptrap_volcano2(
  de_result_1 = res_PACAP$results,
  de_result_2 = res_BDNF$results,
  fdr = FALSE,
  title = "PACAP vs BDNF — Preoptic Area",
  genes.annot = c("Adcyap1", "Bdnf", "Ucn3", "Gng8")
)
```

Genes that sit in the **upper-right quadrant above the diagonal** are more
enriched in PACAP (warm-activated) than BDNF — these are the warm-specific
genes. Genes near the diagonal are similarly enriched in both groups, meaning
they are expressed in the target cell type regardless of thermal state.

### Customising colours

The three DE classes are named after the treatment labels in your data. Pass a
`colors` vector using those names:

```{r volcano2-custom, eval = FALSE}
ptrap_volcano2(
  res_PACAP$results,
  res_BDNF$results,
  fdr = FALSE,
  colors = c(
    "DE in both" = "#D55E00",
    "DE only PACAP" = "#0072B2",
    "DE only BDNF" = "#009E73"
  )
)
```

---

## Providing sample metadata manually

The examples above rely on automatic column-name parsing. If your column names
do not follow the expected pattern, you can supply a `sample_df` data frame
explicitly. It must have at minimum a `sample` column (matching the count
matrix column names), a `fraction` column, and a `block` column:

```{r sample-df, eval = FALSE}
sample_df <- data.frame(
  sample = c("s1_ip", "s1_input", "s2_ip", "s2_input"),
  treatment = c("pb", "pb", "pb", "pb"),
  fraction = c("IP", "INPUT", "IP", "INPUT"),
  block = c("1", "1", "2", "2")
)

res <- ptrap_de(
  counts_mat = counts_mat,
  sample_df = sample_df,
  gene_ids = gene_ids,
  treatment_name = "pb",
  test_method = "LRT"
)
```

---

## Common issues

**"Cannot identify IP/INPUT fraction in column name"**
: The parser could not find an IP or INPUT keyword. Check your column names and
  use `ip_level` / `input_level` to match your exact spelling (e.g.,
  `input_level = "Input"` for mixed case).

**"Cannot identify a replicate number in column name"**
: Every sample column must contain a digit for the replicate number (e.g., `1`,
  `2`, `10`). Add a number to your column names or supply `sample_df` manually.

**FDR column is all `NA` or very few significant genes**
: With n = 2, multiple-testing correction is extremely conservative. Try
  `fdr = FALSE` in the volcano functions. With ≥ 4 replicates and a GLM
  method, FDR-based calls are more reliable.

**`norm.method = "RPKM"` requires `gene.length`**
: Gene lengths (in base pairs) must be provided as a named numeric vector whose
  names match the gene IDs in your count matrix. These are available from
  annotation packages such as `TxDb.*` or via BioMart. If you already have a
  pre-normalised RPKM matrix, use `norm.method = "none"` instead — no lengths
  needed.

---

## Session info

```{r session-info}
sessionInfo()
```
