---
title: "Getting Started with pTRAPPING"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with pTRAPPING}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "90%",
  dpi = 150,
  message = FALSE,
  warning = FALSE
)
```

## What problem does pTRAPPING solve?

Imagine you want to know which genes are actively being read — translated into
protein — in a specific, rare cell type buried inside a complex tissue like the
brain. You cannot easily sort those cells out; they are intermingled with
hundreds of other cell types. TRAP-seq and PhosphoTRAP are molecular techniques
that let you tag ribosomes (the cellular machinery that reads genes) in your
target cells, and then pull them out with their attached RNA. The RNA you
capture in this immunoprecipitation step — the **IP fraction** — reflects what
your target cells were actively translating. You keep the leftover, un-pulled
RNA too — the **INPUT fraction** — as a reference for total gene expression in
the tissue.

The core question is then: **which genes are enriched in the IP relative to the
input?** A gene enriched in IP is being translated in your target population; a
gene depleted in IP is present in the tissue but not in your cells of interest.
Answering this question requires normalising the data, picking a statistical
test, adjusting for multiple comparisons across thousands of genes, and
visualising the results — all of which pTRAPPING handles for you with three
functions.

**PhosphoTRAP** adds one more layer: instead of constitutively tagging
ribosomes, it captures only ribosomes that carry a phosphorylated form of the
S6 protein — a marker of recent neural activation. This means you can ask not
just "which genes are translated in these cells?" but "which genes are
translated in these cells *right now*, after a specific stimulus?". The
experimental design is otherwise identical: IP fraction (activated cells'
RNA) vs. INPUT fraction (total tissue RNA).

## The example datasets: Tan et al. (2016)

Throughout this vignette we use data from
[Tan et al. (2016) *Cell* 167, 47–59](https://doi.org/10.1016/j.cell.2016.08.028),
a landmark PhosphoTRAP study in mice.

**Dataset 1: Identify warm-sentive neurons** 
**_Experiment:_** To identify which genes are highly enriched in warm-sensitive neurons, mice were challenged with heat and phosphorilated ribosomes in neurons that fired in response were isolated by
immunoprecipitation. RNA from immunoprecipitates was sequenced and fold-enrichment for each gene was calculated as IP/Input fractions. This data set has RPKM normalized counts (counts per million, scaled by TMM-adjusted library sizes, and divided by gene length in kilobases) for 8863 genes across 3 biological replicates (3 mice - IP and Input sample each).

```{r load-dataset 1}
library(pTRAPPING)
library(dplyr)
library(tibble)

# Pre-computed RPKM values — already normalised; use with norm.method = "none"
warm_counts <- read.delim(
  system.file("extdata", "warm_counts.txt", package = "pTRAPPING")
)

dim(warm_counts) # 8863 genes × 9 columns (1 gene-ID + 8 samples)
names(warm_counts) # column names encode treatment, fraction, and replicate
```


The column names follow the convention pTRAPPING expects: each name contains a
**treatment label** (`WC - warm-cells`), a **fraction keyword** (`IP` or
`Input`), and a **replicate number** (`1` or `2`), in any order and separated
by underscores. The function parses these automatically — no metadata table
needed.

---

## Step 1 — Differential expression with `ptrap_de()`

`ptrap_de()` compares the IP fraction to the INPUT fraction for a given
treatment group and returns a ranked table of genes based on four available methods. Use this table as a quick guide to pick you model and normalisation method:

| Feature | DESeq2 | edgeR LRT | edgeR QLF | limma/voom |
|---|---|---|---|---|
| **Normalization** | Median-of-ratios (geometric) | TMM | TMM | Quantile normalization |
| **Core assumption** | Most genes not DE | Most genes not DE | Most genes not DE | No DE assumption; matches distributions |
| **How it works** | Per-gene ratio to geometric mean → median ratio = size factor | Weighted mean of log-ratios vs. reference; trims extreme genes | Same as LRT | Forces identical count distributions across all samples |
| **What is scaled** | Raw counts ÷ size factor | Library sizes (indirect) | Library sizes (indirect) | Full distribution remapped per sample |
| **Statistical model** | Negative binomial GLM + dispersion shrinkage | NB GLM; likelihood ratio test | NB GLM; quasi-likelihood F-test | Linear model; `voom` precision weights handle count variance |
| **Test statistic** | Wald test (default) or LRT | Chi-squared LRT | F-test with empirical Bayes squeezing | Moderated t/F-test (empirical Bayes) |
| **Dispersion estimation** | Shared + gene-wise + MAP shrinkage | Common → trended → tagwise | Common → trended → tagwise + QL dispersion | Mean-variance trend via `voom` |
| **Small sample behavior** | Good; shrinkage stabilizes estimates | Conservative; can be anti-conservative with few replicates | Better than LRT with few replicates; QL adds extra variance component | Good; empirical Bayes borrows strength across genes |
| **Key functions** | `estimateSizeFactors()`, `DESeq()` | `calcNormFactors()`, `glmFit()`, `glmLRT()` | `calcNormFactors()`, `glmQLFit()`, `glmQLFTest()` | `normalizeQuantiles()`, `voom()`, `lmFit()` |
| **Data type** | RNA-seq counts | RNA-seq counts | RNA-seq counts | Microarrays (native); RNA-seq via `voom` |
| **Strengths** | Small *n*; LFC shrinkage | Well-established; fast | More reliable with small *n*; recommended over LRT for most RNA-seq | Large datasets; microarrays; complex linear models |
| **Bioconductor pkg** | `DESeq2` | `edgeR` | `edgeR` | `limma` |

All three GLM-based methods (`"LRT"`, `"QLF"`, `"deseq"`) and the LM-based `"voom"` normalise
the data internally — you do not need to normalise first. For the t-test
methods you can choose a normalisation via `norm.method`:

| `norm.method` | When to use |
|---|---|
| `"CPM"` (default) | Raw counts; comparable across replicates of the same gene |
| `"RPKM"` | Raw counts + gene lengths available via `gene.length`; needed for cross-gene comparisons within a sample |
| `"mratios"` | Raw counts; DESeq2-style size-factor normalisation |
| `"none"` | Data already normalised (e.g., a pre-computed RPKM matrix) |

> **Note on `"RPKM"`:** this option requires you to supply gene lengths via
> `gene.length` (a named numeric vector, in base pairs). Without it, the
> function will throw an error. When you already have a
> pre-normalised matrix — as in the Tan et al. dataset below — use
> `norm.method = "none"` instead.
>
> **Note on `filter = FALSE`:** by default, `ptrap_de()` removes low-expression
> genes using `filterByExpr`, which expects raw integer counts. When passing a
> pre-normalised matrix (RPKM, TPM, or similar), the small floating-point values
> will mislead the filter and cause real genes to be dropped. Set `filter = FALSE`
> whenever `norm.method = "none"` is used with a pre-normalised input.

### Calculating gene fold enrichment and p-values with the paired t-test

We use the paired
t-test — the same approach as the original paper. Tan et al. ran the test on
RPKM values, so we pass the pre-normalised RPKM matrix and set
`norm.method = "none"` and `filter = FALSE` to skip any additional normalisation step. Also, pior.count is set to 0 because RPKM values are already normalised and do not require a pseudocount.

```{r ptrap-de-ttest}
warm_de <- warm_counts |>
  dplyr::filter(dplyr::if_any(dplyr::where(is.numeric), ~ .x > 1)) |> #keep counts > 1
  ptrap_de(
    treatment_name = "WC", # only one treatment group in this dataset
    test_method = "paired.ttest",
    norm.method = "none", # data are pre-normalised
    filter = FALSE, # skip count-based filtering
    prior.count = 0 # no pseudocount needed for RPKM values
  )
```

The function prints a message showing what it auto-parsed from the column
names: treatments, replicate numbers (blocks), and fractions. The argument "paired.ttest" in the test_method argument (as well as "unpaired.ttest") returns a named list including `$results` and `$fe` tables. Inspect the result:

```{r inspect-results}
# The paired.ttest method always returns a named list
names(warm_de)

# $results: one row per gene, sorted by p-value
head(warm_de$results)
```

### `$results`: per-gene differential expression. 

The `$results` output is a tibble with one row per gene, sorted by p-value. The columns include log fold change (logFC), p-value, FDR, and a classification of whether the gene is differentially expressed (DE) based on user-defined thresholds for logFC and FDR. By default, the function uses a logFC threshold of 1 (meaning a 2-fold change) and an FDR threshold of 0.05 to classify genes as "UP" (enriched in IP), "DOWN" (depleted in IP), or "NO" (not DE).


### Understanding the output columns

| Column | What it means |
|---|---|
| `Gene` | Gene identifier |
| `logFC` | log2(IP / INPUT) — positive = enriched in target cells |
| `PValue` | Raw p-value from the paired t-test |
| `FDR` | Benjamini-Hochberg adjusted p-value (false discovery rate) |
| `diffexpressed` | `"UP"` (enriched), `"DOWN"` (depleted), or `"NO"` |

The `diffexpressed` column is determined by two thresholds (both must be met):

* `lfc_threshold` — minimum absolute logFC (default `1`, meaning ≥ 2× change)
* `fdr_threshold` — maximum FDR (default `0.05`)

You can change these without re-running the analysis — `ptrap_volcano()` and
`ptrap_volcano2()` recompute the classification from your chosen thresholds.

### `$fe`: per-animal fold enrichments

The paired t-test also returns a `$fe` tibble with each animal's IP/INPUT fold
enrichment (FE) for every gene. This is useful for spot-checking individual
genes or running your own downstream analyses.

```{r inspect-fe}
# Per-animal fold enrichment for the top 5 genes
head(warm_de$fe, 5)
```

## Step 2 — Visualise the results with `ptrap_volcano()`

A **volcano plot** puts logFC on the x-axis and statistical significance
(-log p-value) on the y-axis. Genes in the **upper-right corner** are both
strongly enriched *and* statistically significant — the most interesting
candidates. Genes in the upper-left are strongly depleted. Grey points fail
one or both thresholds. The argument `genes.annot` takes a character vector of gene names to label on the plot. In the example below, we label highly enriched and known warm-sensitive markers shown in figure 1D in Tan et al. (2016), including Adcyap1 (PACAP) and Bdnf. 

```{r ptrap-volcano, fig.cap = "Volcano plot showing known warm-sensitive markers in Tan et al. (2016)."}
warm_de$results |>
  ptrap_volcano(
    fdr = FALSE, # use raw p-values (n=3; FDR is very conservative)
    log_base = 2, # -log2(p) on y-axis, matching Tan et al. 2016
    genes.annot = c(
      "Fosl2",
      "Egr2",
      "Bdnf",
      "Adcyap1",
      "Junb",
      "Fosb",
      "Rrad",
      "Gadd45b"
    ),
    point_alpha = 0.3,
    title = "Differential gene expression in warm-sensitive neurons"
  )
```

Key arguments:

| Argument | Default | What it does |
|---|---|---|
| `fdr` | `TRUE` | Use FDR (`TRUE`) or raw p-value (`FALSE`) for the y-axis and colouring |
| `lfc_threshold` | `1` | logFC cutoff for the vertical dashed lines |
| `fdr_threshold` | `0.05` | p-value cutoff for the horizontal dashed line |
| `log_base` | `10` | Base of the log transform on the y-axis |
| `genes.annot` | `NULL` | Character vector of gene names to label |
| `title` | auto | Plot title (defaults to region + treatment) |

### Customising colours

Pass a named vector to override the defaults:

```{r volcano-custom, eval = FALSE}
warm_de$results |>
  ptrap_volcano(
    fdr = FALSE,
    colors = c("UP" = "#D55E00", "DOWN" = "#56B4E9"),
    title = "Volcano plot — custom colours"
  )
```

The function returns a ggplot2 object, so you can add or override any layer
with standard ggplot2 syntax:

```{r volcano-ggplot, eval = FALSE}
library(ggplot2)

warm_de$results |>
  ptrap_volcano(fdr = FALSE) +
  theme(plot.title = element_text(face = "bold")) +
  xlim(-10, 10)
```

---

## Step 3 — Compare two conditions with `ptrap_volcano2()`

**Dataset 2: Characterising warm-sensitive neurons** 
**_Experiment:_** The authors created BDNF-Cre and PACAP-Cre transgenic mice and used TRAP to sequence the genes expressed in these warm-sensitive neuron populations. 
This data set has RPKM normalized counts for 8863 genes across 2 biological replicates (2 mice - IP and Input sample each) for both BDNF-Cre and PACAP-Cre lines.  

```{r load-dataset-2}
# Pre-computed RPKM values — already normalised; use with norm.method = "none"
counts_rpkm <- read.delim(
  system.file("extdata", "TAN_etal_2016_RPKM.txt", package = "pTRAPPING")
)

dim(counts_rpkm) # 8863 genes × 9 columns (1 gene-ID + 8 samples)
names(counts_rpkm) # column names encode treatment, fraction, and replicate
```

Run `ptrap_de()` for both conditions separately to get two DE result tables, which are the input for `ptrap_volcano2()`. 

```{r ptrap-de-pacap}
pacap_de <- counts_rpkm |>
  filter(dplyr::if_any(dplyr::where(is.numeric), ~ .x > 1)) |> #keep counts > 1
  mutate(gene = make.unique(Gene)) |>
  column_to_rownames("gene") |>
  #round() |>
  ptrap_de(
    test_method = "paired.ttest",
    norm.method = "none",
    treatment_name = "PACAP",
    filter = FALSE,
    lfc_threshold = 0.3,
    prior.count = 0
  )

# Inspect the per-gene fold enrichment
pacap_de$fe

bdnf_de <- counts_rpkm |>
  filter(dplyr::if_any(dplyr::where(is.numeric), ~ .x > 1)) |> #keep counts > 1
  mutate(gene = make.unique(Gene)) |>
  column_to_rownames("gene") |>
  #round() |>
  ptrap_de(
    test_method = "paired.ttest",
    norm.method = "none",
    treatment_name = "BDNF",
    filter = FALSE,
    lfc_threshold = 0.3,
    prior.count = 0
  )

# Inspect the per-gene fold enrichment
bdnf_de$fe
```

BDNF and PACAP cells co-express both BDNF and PACAP genes and almost the same set of genes, including Unc3, Nxph4, Ghrh, etc.

Then `ptrap_volcano2()` places both
results on a single 2D scatter plot:

* **x-axis**: logFC for condition 2
* **y-axis**: logFC for condition 1
* **Diagonal (dotted)**: genes enriched equally in both
* **Above diagonal**: more enriched in condition 1
* **Below diagonal**: more enriched in condition 2

Genes that meet the significance thresholds in both conditions, in only one,
or in neither are shown in distinct colours.

```{r ptrap-volcano2, fig.cap = "Scatter comparison of PACAP and BDNF enrichment. Genes on the diagonal are enriched equally in both conditions."}
ptrap_volcano2(
  de_result_1 = pacap_de$results,
  de_result_2 = bdnf_de$results,
  fdr = FALSE,
  title = "PACAP vs BDNF — Preoptic Area",
  genes.annot = c(
    "Adcyap1",
    "Bdnf",
    "Ucn3",
    "Gng8",
    "Nxph4",
    "Ghrh",
    "Emx2",
    "Fezf1",
    "Nhlh2"
  )
)
```

Genes near the diagonal are similarly enriched in both PACAP and BDNF cells.

### Customising colours

The three DE classes are named after the treatment labels in your data. Pass a
`colors` vector using those names:

```{r volcano2-custom, eval = FALSE}
ptrap_volcano2(
  pacap_de$results,
  bdnf_de$results,
  fdr = FALSE,
  colors = c(
    "DE in both" = "#D55E00",
    "DE only PACAP" = "#0072B2",
    "DE only BDNF" = "#009E73"
  )
)
```

---

## Common issues

**"Cannot identify IP/INPUT fraction in column name"**
: The parser could not find an IP or INPUT keyword. Check your column names and
  use `ip_level` / `input_level` to match your exact spelling (e.g.,
  `input_level = "Input"` for mixed case).

**"Cannot identify a replicate number in column name"**
: Every sample column must contain a digit for the replicate number (e.g., `1`,
  `2`, `10`). Add a number to your column names or supply `sample_df` manually.

**FDR column is all `NA` or very few significant genes**
: With n = 2, multiple-testing correction is extremely conservative. Try
  `fdr = FALSE` in the volcano functions. With ≥ 4 replicates and a GLM
  method, FDR-based calls are more reliable.

**`norm.method = "RPKM"` requires `gene.length`**
: Gene lengths (in base pairs) must be provided as a named numeric vector whose
  names match the gene IDs in your count matrix. These are available from
  annotation packages such as `TxDb.*` or via BioMart. If you already have a
  pre-normalised RPKM matrix, use `norm.method = "none"` instead — no lengths
  needed.

---